Colors and Residual-based Shadings in the Strucplot Framework

### 2005-10-02 trunk/vcd/inst/doc/shading.Rnw
Vignettes overworked.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@466 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}

@@

%\VignetteIndexEntry{Colors and residual-based Shadings in Strucplots}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

@@

\title{Colors and Residual-based Shadings in Strucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\email{David.Meyer@R-project.org}
}
\maketitle
\sloppy

Unlike other graphics functions in base {\sf R}, the strucplot framework 
allows almost full control over the graphical parameters of all plot elements. In
particular, the user can \emph{individually} modify the graphical appearance of  
tiles in association plots, mosaic plots, and sieve plots. 
Built on top of this functionality, the
framework supplies a set of shading functions choosing colors
appropriate for the visualization of log-linear models.
The tiles' graphical parameters are set using the \code{gp} argument
of the strucplot functions. This argument basically expects an object
of class \code{gpar} whose components are arrays of the same shape 
(length and dimensionality) than the data table (see Section \ref{sec:gp}). 
For convenience, however, the user can also
supply a function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
parameters and returns such a function (see Section
\ref{sec:shading}). We provide several shading functions, including
support for both the HSV and the HCL color space, and the
visualization of significance tests (see Section
\ref{sec:overview}).

@@

As an example, consider the \data{UCBAdmissions} data. 
In the table collapsed over departments, we
would like to highlight the (incidentally wrong) impression that there
were to much male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

@@

\noindent As the example shows, we create a fourfold table 
with appropriate colors (dark cyan for admitted male students and dark
magenta for rejected female students) and supply them to the \code{fill} component 
of the \code{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
For optical clarity, we addtionally hide the tiles' borders by setting the \code{col} 
component to 0 (white).

If the parameters specified in the \code{gpar} object are incomplete,
strucplot functions will recycle them along the last splitting
dimension. In the following example based on the \data{Titanic} data, 
we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

@@

\noindent Note that \codefun{spacing\_highlighting} sets the spaces
between tiles in the last dimension to 0. The \code{labeling\_args} 
argument ensures that labels do not overlapp (see the separate
vignette on labeling for more information).

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of the observed from the expected
values under a given log-linear model. 
The idea is to visualize at least the size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

@@

\begin{enumerate}
\item Pre-computing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \code{"gpar"} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in 1.
\item Providing a \emph{generating} function taking parameters and returning
  a function as described in 2. The parameters of this generating
  function are either directly specified (which is not possible for
  all shading functions), or as a list to the \code{shading\_args} argument.
\end{enumerate}

@@

@@

(fillcols <- ifelse(residuals > 0, "lightblue", "lightsalmon"))

mosaic(ucb, gp = gpar(fill = fillcols))
@

@@

For implementing method 2, we need to create a `shading function'
that---technically speaking---computes \code{"gpar"} objects from
residuals. For that, we can just reuse the code from the previous step:

<<shading2>>=
shade_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
@

\noindent To create a mosaic display with binary shading, 
it now suffices to specify the data table along with
\codefun{shade\_fun}:

<<shading3>>=
mosaic(ucb, gp = shade_fun)
@

\noindent \codefun{mosaic} internally calls
\codefun{strucplot} which computes the residuals from the specified
independence model (total independence by default), passes them to
\codefun{shade\_fun}, and uses the \code{"gpar"} object returned 
to finally create the plot.

Our \codefun{shade\_fun} function might be useful, 
but suffers from a major flaw: the colors are hard-wired. We cannot
simply extend the argument list to include, e.g., 
a \code{fill = c("lightblue", "lightsalmon")} argument because
\codefun{strucplot} will neither know how to handle it, nor let us
change the defaults. In fact, the interface of shading functions is fixed,
they are expected to take exactly one argument: a table of residuals. 
This is where generating functions (method 3) come
into play. We simply wrap our shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. This kind of wrapping function
is also called a \emph{generating function} since it returns (`creates') a
`parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The follwing example shows the necessary
extensions for our running example:

<<shading3>>=
myshading_binary <- function(col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
@

\noindent In the call of \codefun{mosaic} (figure not shown), 
using the new \codefun{myshading\_binary} function, we can now simply change the colors:

<<shading4>>=
mosaic(ucb, gp = myshading_binary(c("red","blue")))
@

\noindent The procedure described so far is a rather general concept,
applicable to a wide family of user-level \pkg{grid} graphics. Indeed, the
customization of other components of the strucplot framework
(labeling, spacing, legend, and core functions) follows the same idea.
Now for the shading functions, this is not the whole story. Since shading
functions can be used for visualizing significance (see next section), it
makes sense for generating functions to have access to the model,
i.e., observed and expected values, residuals, and degrees of
freedom. For example, the \codefun{shading\_max} generating function
computes a permutation distribution of the maximum statistic and $p$
values for specified significance levels 
based on the observed table to create data-driven cut-off
points. If this was done in the shading function itself, the
permutation statistic would be recomputed every time the shading
function is called, resulting in possibly severe performance loss 
and numerical inconsistencies. Therefore, generating functions for
shadings are required to take at least the parameters 
\code{observed}, \code{expected}, \code{residuals}, and \code{df}
(these are provided by the strucplot framework),
followed by other parameters controlling the shading
appearance (to be specified by the user):

<<shading4>>=
myshading_binary2 <- function(observed = NULL, residuals = NULL, expected = NULL, 
    df = NULL, col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
class(myshading_binary2) <- "generating_function"
@

In some sence, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{myshading\_binary2}) this is optional, but
recommended for consistency:

<<shading5>>=
mosaic(ucb, gp = myshading_binary2, gp_args = list(c("red","blue")))
@

@@

Apart from \codefun{shading\_binary}, there are currently two basic shadings available:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using the HSV space, and
then briefly explain the differences to the HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
The HCL space is trickier
to use, but preferable to the HSV space from a perceptual point of view.

In the HSV space, colors are specified in three dimensions: Hue,
Saturation (`colorfulness'), and Value (`brightness', amount of gray). 
These three dimensions are used by \codefun{shading\_hsv} to visualize information 
about the residuals and the underlying independence model. The hue 
indicates the residuals' sign: by default, blue for positive, and red
for negative residuals. The saturation of a residual 
is set according to its size: high saturation for large, and low
saturation for small residuals. 
Finally, the overall brightness is used to indicate the significance of
a test statistic: bright colors for significant, 
and dark colors for non-significant results.

@@

\begin{figure}[h]
\begin{center}
<<haireyefig,fig=TRUE,echo=FALSE>>=
<<haireye>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent Large positive residuals (greater than $4$) can be found for
brown eyes/black hair and blue eyes/blond hair, they are colored in saturated red. 
On the other hand, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep blue. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
are shaded in white.
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ level, respectively. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure \ref{fig:interpolate}):

@@

\noindent The \code{interpolate} argument also accepts a function,
which then is called with the absolute residuals to get a vector of
cut-off points. Thus, it is possible to automatically choose the cut-off points 
in a data-driven way. For example,
one might think that the extension from four cut-off points 
to a continuous shading---visualizing the whole range of 
significance levels---could be useful. We simply need a one-to-one
mapping from the residuals to the $p$ values:

@@

\noindent Note that this \codefun{ipol} 
function maps residuals greater than 4 to a significance
level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
deceiving:

<<continuous2>>=
mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
labeling_args = list(abbreviate = c(Sex = 3)))
@

@@

The third remaining dimension, the value, is used for
visualizing the significance of a test statistic. The user can either
directly specify the $p$ value, or, alternatively, a function that
computes it, to the \code{p.value} argument. Such a function must take
observed and expected values, residuals, and degrees of freedom as arguments. 
If nothing is specified, the $p$ value is computed from 
a $\chi^2$ distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
otherwise dark colors are employed. 
The following example using the \data{Bundesliga} data 
shows the relationship of home goals and away
goals of the German premier's soccer league in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
$\chi^2$ test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

@@

A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
miming the shading introduced by \cite{vcd:Friendly:1994}. In addition
to the defaults of the HSV shading, it uses the border color and line type to 
redundantly code the residuals' sign. The following example again uses 
the \data{Bundesliga} data from above, this time using the Friendly 
scheme and, in addition, an alternative legend (see Figure \ref{fig:friendly}):

@@

\noindent (The \code{zeroSize = 0} argument removes the bullets indicating zero 
observed values. This feature does not
exist in the original {\sf SAS} implementation of the Friendly mosaic plots.)

As introduced before, the default shading scheme is not
\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
perceptual characteristics of the HCL color space.
Figure \ref{fig:shadingHSVHCL} depicts color scales for red and blue hue
(left and right side, respectively) in the HSV and HCL spaces (upper and lower part, respectively), varying colorfulness and brightness ($x$ and $y$-axis, respectively).
The boxes represent the diverging color palettes used for the shadings.
For the HSV space, we can can see that the effect of changing the 
level of brightness (`value') is not the same for different levels of 
saturation, and again not the same for the two different hues.
In fact, in the HSV space all dimensions are confounded, which 
obviously is problematic for coding information. In contrast, the HCL color
space offers perceptual uniform colors: as can be seen from the lower part, 
the chroma is homogeneous for different levels of luminance. 
Unfortunately, this comes at the 
price of the space being irregularly shaped, making it difficult to automatically select 
diverging color palettes. The following example again 
illustrates the \data{HairEyeColor} data, this time with HCL colors 
(Figure \ref{fig:hclshading1} depicts the default palette, 
and Figure \ref{fig:hclshading2} an alternative setting):

@@

\noindent The maximum test is significant although the residuals are
all in the $[-2,2]$ interval, as the cut-off points are even smaller.
The \codefun{shading\_hcl} function with default cut-off points would
not have shown any color. In addition, since the test statistic is the
maximum of the absolute Pearson residuals, \emph{every} colored
residual violates the null hypotheses of independence, and thus, the
`culprits' can immediately be identified.

\bibliographystyle{apalike}
\bibliography{vcd}

>>>>

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\class}[1]{\textsf{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}

@@

%\VignetteIndexEntry{Colors and Residual-based Shadings in the Strucplot Framework}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

@@

\title{Colors and Residual-based Shadings in the Strucplot Framework}
\author{by David Meyer, Achim Zeileis, and Kurt Hornik}
\date{}
\maketitle
\sloppy

\section{Introduction}

Unlike other graphics functions in base {\sf R}, the strucplot framework 
allows almost full control over the graphical parameters of all plot elements. In
particular, in association plots, mosaic plots, and sieve plots, 
the user can modify the graphical appearance of each tile individually. 
Built on top of this functionality, the
framework supplies a set of shading functions choosing colors
appropriate for the visualization of log-linear models.
The tiles' graphical parameters are set using the \code{gp} argument
of the functions of the strucplot framework. This argument basically expects an object
of class \class{gpar} whose components are arrays of the same shape 
(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
For convenience, however, the user can also
supply a specialized graphical appearance control (``grapcon'') 
function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
certain arguments and returns such a grapcon function (see Section
\ref{sec:shading}). We provide several shading functions, including
support for both HSV and HCL colors, and the
visualization of significance tests (see Section \ref{sec:overview}).

@@

As an example, consider the \data{UCBAdmissions} data. 
In the table aggregated over departments, we
would like to highlight the (incidentally wrong) impression that there
were too many male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

@@

\noindent As the example shows, we create a fourfold table 
with appropriate colors (dark cyan for admitted male students and dark
magenta for rejected female students) and supply them to the \code{fill} component 
of the \class{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
For visual clarity, we additionally hide the tiles' borders by setting the \code{col} 
component to 0 (white).

If the parameters specified in the \class{gpar} object are ``incomplete'',
they will be recycled along the last splitting
dimension. In the following example based on the \data{Titanic} data, 
we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

@@

\noindent Note that \codefun{spacing\_highlighting} sets the spaces
between tiles in the last dimension to 0. The \code{labeling\_args} 
argument ensures that labels do not overlap (see the separate
vignette: ``Labeling in the Strucplot Framework'' for more information).

@@

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
values under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

@@

\begin{enumerate}
\item Pre-computing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
\item Providing a grapcon \emph{generating} function taking parameters and returning
  a function as described in alternaive 2. 
\end{enumerate}

@@

\subsubsection*{Alternative 1: precomputed \class{gpar} object}

@@

(shading1_obj <- ifelse(residuals > 0, "lightblue", "lightsalmon"))

mosaic(ucb, gp = gpar(fill = shading1_obj))
@

@@

\subsubsection*{Alternative 2: grapcon function}

For implementing alternative 2, we need to create a `shading function'
that computes \class{gpar} objects from
residuals. For that, we can just reuse the code from the previous step:

<<shading2>>=
shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
class(shading2_fun) <- c("grapcon","shading")
@

\noindent Note the \code{class} attribute set on \codefun{shading2\_fun}.
To create a mosaic display with binary shading, 
it now suffices to specify the data table along with
\codefun{shading2\_fun}:

<<shading3>>=
mosaic(ucb, gp = shading2_fun)
@

\noindent \codefun{mosaic} internally calls
\codefun{strucplot} which computes the residuals from the specified
independence model (total independence by default), passes them to
\codefun{shading2\_fun}, and uses the \class{gpar} object returned 
to finally create the plot.

Our \codefun{shading2\_fun} function might be useful, 
but can still be improved: the hard-wired colors should be
customizable. We cannot
simply extend the argument list to include, e.g., 
a \code{fill = c("lightblue", "lightsalmon")} argument because
\codefun{strucplot} will neither know how to handle it, nor let us
change the defaults. In fact, the interface of shading functions is fixed,
they are expected to take exactly one argument: a table of residuals. 
This is where generating functions (alternative 3) come
into play.

\subsubsection*{Alternative 3: grapcon-generating function}

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The follwing example shows the necessary
extensions for our running example:

<<shading3>>=
shading3a_fun <- function(col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
  structure(ret, class = c("grapcon","shading"))
}
@

\noindent In the call of \codefun{mosaic} (figure not shown), 
using the new \codefun{shading3a\_fun} function, we can now simply change the colors:

<<shading4>>=
mosaic(ucb, gp = shading3a_fun(c("red","blue")))
@

\noindent The procedure described so far is a rather general concept,
applicable to a wide family of user-level \pkg{grid} graphics. Indeed, the
customization of other components of the strucplot framework
(labeling, spacing, legend, and core functions) follows the same idea.
Now for the shading functions, more customization is needed. Note that
\codefun{shading3a\_fun} needs to be evaluated by the user, even if the
defaults shall be used. It is a better idea to let \codefun{strucplot} call
the generating function, which, in particular, allows the passing of arguments that
are computed by \codefun{strucplot}. Since shading
functions can be used for visualizing significance (see Section 
\ref{sec:overview}), it makes sense for generating functions to have access to the model,
i.e., observed and expected values, residuals, and degrees of
freedom. For example, the \codefun{shading\_max} generating function
computes a permutation distribution of the maximum statistic and $p$
values for specified significance levels 
based on the observed table to create data-driven cut-off
points. If this was done in the shading function itself, the
permutation statistic would be recomputed every time the shading
function is called, resulting in possibly severe performance loss 
and numerical inconsistencies. Therefore, generating functions for
shadings are required to take at least the parameters 
\code{observed}, \code{expected}, \code{residuals}, and \code{df}
(these are provided by the strucplot framework),
followed by other parameters controlling the shading
appearance (to be specified by the user):

<<shading4>>=
shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
    df = NULL, col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
  structure(ret, class = c("grapcon","shading"))
}
class(shading3b_fun) <- c("grapcon_generator","shading")
@

In some sence, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:

<<shading5>>=
mosaic(ucb, gp = shading3b_fun, gp_args = list(col = c("red","blue")))
@

@@

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using the HSV space, and
then briefly explain the differences to the HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
The HCL space is trickier
to use, but preferable to the HSV space from a perceptual point of view.

In the HSV space, colors are specified in three dimensions: Hue,
Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
These three dimensions are used by \codefun{shading\_hsv} to visualize information 
about the residuals and the underlying independence model. The hue 
indicates the residuals' sign: by default, blue for positive, and red
for negative residuals. The saturation of a residual 
is set according to its size: high saturation for large, and low
saturation for small residuals. 
Finally, the overall lightness is used to indicate the significance of
a test statistic: light colors for significant, 
and dark colors for non-significant results.

@@

\begin{figure}[h]
\begin{center}
<<haireyefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<haireye>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent Large positive residuals (greater than $4$) can be found for
brown eyes/black hair and blue eyes/blond hair, and are colored in saturated blue. 
On the other hand, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
are shaded in white.
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure \ref{fig:interpolate}):

@@

\noindent The elements of the numeric vector passed to
\code{interpolate} define the knots of an interpolating step function
used to map the absolute residuals to saturation levels. 
The \code{interpolate} argument also accepts a user-defined function,
which then is called with the absolute residuals to get a vector of
cut-off points. Thus, it is possible to automatically choose the cut-off points 
in a data-driven way. For example,
one might think that the extension from four cut-off points 
to a continuous shading---visualizing the whole range of 
residuals---could be useful. We simply need a one-to-one
mapping from the residuals to the saturation values:

@@

\noindent Note that this \codefun{ipol} 
function maps residuals greater than 4 to a saturation
level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
deceiving:

<<continuous2>>=
mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
labeling_args = list(abbreviate = c(Sex = TRUE)))
@

@@

The third remaining dimension, the value, is used for
visualizing the significance of a test statistic. The user can either
directly specify the $p$ value, or, alternatively, a function that
computes it, to the \code{p.value} argument. Such a function must take
observed and expected values, residuals, and degrees of freedom (used
by the independence model) as arguments. 
If nothing is specified, the $p$ value is computed from 
a $\chi^2$ distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, light colors are used, 
otherwise dark colors are employed. 
The following example using the \data{Bundesliga} data 
shows the relationship of home goals and away
goals of Germany's premier soccer league in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
$\chi^2$ test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

@@

A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
implementing the shading introduced by \cite{vcd:Friendly:1994}. In addition
to the defaults of the HSV shading, it uses the border color and line type to 
redundantly code the residuals' sign. The following example again uses 
the \data{Bundesliga} data from above, this time using the Friendly 
scheme and, in addition, an alternative legend (see Figure \ref{fig:friendly}):

@@

\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
observed values. This feature is not provided
in the original {\sf SAS} implementation of the Friendly mosaic plots.)

As introduced before, the default shading scheme is not
\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
perceptual characteristics of the HCL color space.
Figure \ref{fig:shadingHSVHCL} depicts 
the HSV space in the upper panel and the HCL space in the lower panel.
On the left (right) side, we see the color scales for red (blue)
hue, respectively. The $x$-axis represents the colorfulness, and the
$y$-axis the brightness.
The boxes represent the diverging color palettes used for the shadings.
For the HSV space, we can see that the effect of changing the 
level of brightness (`value') is not the same for different levels of 
saturation, and again not the same for the two different hues.
In fact, in the HSV space all dimensions are confounded, which 
obviously is problematic for coding information. In contrast, the HCL color
space offers perceptually uniform colors: as can be seen from the lower panel, 
the chroma is homogeneous for different levels of luminance. 
Unfortunately, this comes at the 
price of the space being irregularly shaped, making it difficult to automatically select 
diverging color palettes. The following example again 
illustrates the \data{HairEyeColor} data, this time with HCL colors 
(Figure \ref{fig:hclshading1} depicts the default palette, 
and Figure \ref{fig:hclshading2} an alternative setting):

@@

\noindent The maximum test is significant although the residuals are
all in the $\left[-2,2\right]$ interval.
The \codefun{shading\_hcl} function with default cut-off points would
not have shown any color. In addition, since the test statistic is the
maximum of the absolute Pearson residuals, \emph{each} colored
residual violates the null hypotheses of independence, and thus, the
`culprits' can immediately be identified.

\bibliographystyle{jss}
\bibliography{vcd}
